\vspace{-1.5em} \bf A general and efficient representation of ancestral recombination graphs

### 2023-10-12 paper.tex
Yans updates (applying comments from Andrew)

\begin{figure}
\begin{center}
    \includegraphics[width=\textwidth]{illustrations/arg-in-pedigree}
\end{center}
\caption{\label{fig-arg-in-pedigree}
% TODO caption needs another pass. One the full section has been written
% it'll be clearer what we need to say in the caption vs the text.
An example genome ARG embedded in a pedigree.
(A) Diploid individuals, visualised in a highly inbred pedigree and
labelled $D_1$ to $D_8$,
contain both paternal and maternal  genomes
labelled \textsf{a} to \textsf{p}. Black lines show inheritance paths connecting
genomes in the current generation (\textsf{a} to \textsf{d}) with their ancestors.
Genomes \textsf{a} and \textsf{c} are the product of two independent
recombination events between
the paternal genomes \textsf{e}
and \textsf{f} (with breakpoints in red), and regions of genome inherited are shown.
Genomes are shaded such that where, backwards in time,
they merge into a common ancestor, the merged region is darker.
(B) The corresponding gARG along with inheritance annotations on all edges
(partial inheritance in bold).
(C) The corresponding local trees.
}
\end{figure}

As illustrated in Fig.~\ref{fig-arg-in-pedigree},
the gARG for a given set of individuals is embedded in their pedigree.
In this diploid example, each genome is inherited from one
of the individual's parents,
and may be the recombined product of that parent's two genomes.
For example, diploid individual $D_1$ in Fig.~\ref{fig-arg-in-pedigree}
has two genomes \noderef{a} and \noderef{b},
inherited from parents $D_3$ and $D_4$.
Genome \noderef{b} was inherited directly from $D_4$'s genome \noderef{g} without
recombination, whereas
\noderef{a} is the recombinant product of
$D_2$'s two genomes \noderef{e} and \noderef{f} crossing over at position 2.
This recombination means \noderef{a} inherits the (half-closed) genome
interval $[0, 2)$ from \noderef{e} and $[2, 10)$ from \noderef{f}.
These intervals are shown attached to the corresponding graph edges.

@@

Whether we regard nodes as events or genomes is something of a
philosophical question, and in most senses the two are equivalent
(but see the discussion on ``unary nodes'' in section XXX).
From a concrete data-structure encoding perspective, the only real
difference between an eARG and a gARG is that we store recombination
\emph{breakpoints} at nodes of out-degree two (recombination nodes)
in an eARG and we store \emph{inheritance intervals} on the edges
of a gARG. A key advantage of associating inheritance intervals with
edges in a gARG is that it allows us to describe evolutionary events other
than simple crossovers, such as multiple recombinations and
gene-conversions. By explicitly stating the exact regions of genome
that are transmitted along an edge, we generalise the representation
to encompass any process that leads to the differential inheritance of
genetic material along the genome.

@@

It is useful to have
a concrete definition and explicit process for local tree extraction.
We can describe a local tree as an
oriented tree~\cite[p.\ 461]{knuth11combinatorial},
a sequence of integers $\pi_1\pi_2\dots$ where $\pi_u$ denotes the parent of
node $u$~\citep{kelleher2013coalescent,kelleher2014coalescent,kelleher2016efficient}.
If $u$ is a root
% I'm not sure if this reachability stuff is a helpful clarification, and whether it'll just
% confuse people. It depends on some tricky details.
(or not reachable from the samples)
in the local tree at position $x$, then $\pi_u = -1$ (by convention).
% JK: this confused some people, so best left out?
% Technically, this structure is a ``forest'', as it allows us to have
% multiple disconnected trees, which occurs when all samples have not
% found a most recent common ancestor.
Oriented trees allow nodes
to have  any number of children; from one (see section XXX for more
discussion on such ``unary nodes'' in the local trees), to arbitrarily
large polytomies.
To recover the local tree $\pi$ at position $x$, we begin by
setting $\pi_u = -1$ for each $u\in N$. Then, for each sample
node in $S$ we trace its path rootwards through the
ARG for position $x$, and record this path in $\pi$.
Specifically, at a given node $u$,
we an find edge $(c, p, I) \in E$ such that $u = c$ and $x \in I$, and set
$\pi_c \leftarrow p$. We then set $u \leftarrow p$, and repeat
until either $\pi_u \neq -1$ (indicating we have traversed this section
of the ARG already on the path from another sample) or there
is no matching outbound edge (indicating we are at a root).

@@

The internal nodes of trees, however, are usually
considered to be \emph{unlabelled},
because they most often correspond
to anonymous hypothetical ancestors.
Internal ARG nodes likewise correspond to anonymous hypothetical
ancestors, but if we do not label them we lose the fact that
the \emph{same} ancestors persist across multiple trees.
Without labelled internal tree nodes,
we must infer their identity
from the tree topology, which is both
expensive to compute and imprecise. For example
we could infer that node \textsf{k} is shared in the second
and third trees of in Fig.~\ref{fig-simplification}D
because it subtends identical subtrees in both,
but it would not be straightforward to infer that it
is also the root of the first tree.

>>>>

\begin{figure}
\begin{center}
    \includegraphics[width=\textwidth]{illustrations/arg-in-pedigree}
\end{center}
\caption{\label{fig-arg-in-pedigree}
% TODO caption needs another pass. One the full section has been written
% it'll be clearer what we need to say in the caption vs the text.
An example genome ARG embedded in a pedigree.
(A) Diploid individuals (blue), visualised in a highly inbred pedigree and
labelled $D_1$ to $D_8$,
contain both paternal and maternal  genomes
labelled \textsf{a} to \textsf{p}. Black lines show inheritance paths connecting
genomes in the current generation (\textsf{a} to \textsf{d}) with their ancestors.
Genomes \textsf{a} and \textsf{c} are the product of two independent
recombination events (red) between
the paternal genomes \textsf{e}
and \textsf{f}, and regions of genome inherited are shown.
Genomes are shaded such that where, backwards in time,
they merge into a common ancestor, the merged region is darker.
(B) The corresponding gARG along with inheritance annotations on all edges
(partial inheritance in bold).
(C) The corresponding local trees.
}
\end{figure}

As illustrated in Fig.~\ref{fig-arg-in-pedigree},
the gARG for a given set of individuals is embedded in their pedigree.
In this figure, diploid individuals are labelled $D_1, D_2...$, and
rather than denoting their constituent genomes using the natural numbers
we use lowercase alphabetical labels, to avoid confusion with genomic
positions. Individual $D_1$ is therefore composed
of two genomes \noderef{a} and \noderef{b}, which are inherited from its
two parents, $D_3$ and $D_4$. Each inherited genome may be the recombined product
of the genomes contained within a parent. In this example,
genome \noderef{b} was inherited directly from $D_4$'s genome \noderef{g} without
recombination, whereas
\noderef{a} is the recombinant product of
$D_2$'s two genomes \noderef{e} and \noderef{f} crossing over at position 2.
This recombination means \noderef{a} inherits the (half-closed) genome
interval $[0, 2)$ from \noderef{e} and $[2, 10)$ from \noderef{f}.
These intervals are shown attached to the corresponding graph edges.

@@

Whether we regard nodes as events or genomes is something of a
philosophical question, and in most senses the two are equivalent
(but see the discussion on ``unary nodes'' in section XXX).
From a concrete data-structure encoding perspective, the only real
difference between an eARG and a gARG is that in an eARG we store recombination
\emph{breakpoints} at nodes of out-degree two (recombination nodes)
whereas in a gARG we store \emph{inheritance intervals} on edges.
A key advantage of associating inheritance intervals with
edges in a gARG is that it allows us to describe evolutionary events other
than simple crossovers, such as multiple recombinations and
gene-conversions. By explicitly stating the exact regions of genome
that are transmitted along an edge, we generalise the representation
to encompass any process that leads to the differential inheritance of
genetic material along the genome.

@@

It is useful to have
a concrete definition and explicit process for local tree extraction.
We can describe a local tree as an
oriented tree~\cite[p.\ 461]{knuth11combinatorial},
a sequence of integers $\pi_1, \pi_2\dots$ where $\pi_u$ denotes the parent of
node $u$~\citep{kelleher2013coalescent,kelleher2014coalescent,kelleher2016efficient}.
If $u$ is a root
% I'm not sure if this reachability stuff is a helpful clarification, and whether it'll just
% confuse people. It depends on some tricky details.
(or not reachable from the samples)
in the local tree at position $x$, then $\pi_u = -1$ (by convention).
% JK: this confused some people, so best left out?
% Technically, this structure is a ``forest'', as it allows us to have
% multiple disconnected trees, which occurs when all samples have not
% found a most recent common ancestor.
Oriented trees allow nodes
to have  any number of children; from one (see section XXX for more
discussion on such ``unary nodes'' in the local trees), to arbitrarily
large polytomies.
To recover the local tree $\pi$ at position $x$, we begin by
setting $\pi_u = -1$ for each $u\in N$. Then, for each sample
node in $S$ we trace its path rootwards through the
ARG for position $x$, and record this path in $\pi$.
Specifically, at a given node $u$,
we find an edge $(c, p, I) \in E$ such that $u = c$ and $x \in I$, and set
$\pi_c \leftarrow p$. We then set $u \leftarrow p$, and repeat
until either $\pi_u \neq -1$ (indicating we have traversed this section
of the ARG already on the path from another sample) or there
is no matching outbound edge (indicating we are at a root).

@@

The internal nodes of trees, however, are usually
considered to be \emph{unlabelled},
because they most often correspond
to anonymous hypothetical ancestors.
Internal ARG nodes likewise correspond to anonymous hypothetical
ancestors, but if we do not label them we lose the fact that
the \emph{same} ancestors persist across multiple trees.
Without labelled internal tree nodes,
we must infer their identity
from the tree topology, which is both
expensive to compute and imprecise. For example
using only the tree topologies it would not be straightforward to
infer that node \noderef{i} in the first and last trees of
Fig.~.\ref{fig-arg-in-pedigree}C is the same node.
